Spatial variation in allometric growth of invasive lionfish has management implications

### 2019-02-11 docs/Manuscript.tex
Updated, ready to resubmit

\begin{abstract}
Lionfish (\textit{Pterois volitans / miles}) are an invasive species in
the Western Atlantic and the Caribbean. Improving management of invasive
lionfish populations requires accurate total biomass estimates, which
depend on accurate estimates of allometric growth, but sedentary species
like lionfish often exhibit high levels of spatial variation in life
history characteristics. We reviewed 17 published length-weight
relationships for lionfish taken throughout their invasive range and
found regional differences that led to significant misestimates when
calculating weight from length observations. The spatial pattern we
observed is consistent with findings from other studies focused on
genetics or length-at-age. We show that the use of \textit{ex situ}
parameters can result in up to a threefold under- or overestimation of
total weight for an individual organism. When estimating total biomass,
the use of other parameters may produce estimates between 76.2\% and
140\% of true observed biomass. These findings can have implications for
management in terms of predicting effects on local ecosystems,
evaluating the effectiveness of removal programs, or estimating biomass
available for harvest.
\end{abstract}

@@

Governments and non-profit organizations have sought to reduce lionfish
densities through removal programs and incentivizing its consumption
\citep{chin_2016}. In some cases, these have shown to significantly
reduce -- but not quite eliminate -- lionfish abundances at local scales
\citep{deleon_2013,sandel_2015}. Complete eradication of lionfish
through fishing is unlikely because of their rapid recovery rates and
ongoing recruitment to shallow-water areas from persistent populations
in mesophotic ecosystems \citep{barbour_2011,andradibrown_2017}.
However, promoting lionfish consumption might create a level of demand
capable of incentivizing a stable fishery while controlling
shallow-water populations, thus creating alternative livelihoods and
avoiding further negative effects to local biota.

The feasibility of establishing fisheries through lionfish removal
programs has been extensively evaluated through field observations and
empirical modeling
\citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,usseglio_2017}.
Determining the feasibility of such initiatives requires modeling the
change in biomass in response to changes in fishing mortality
(\emph{i.e.} culling). A common way to model this is via
length-structured population models, where fish lengths are converted to
weight to calculate total biomass
\citep{barbour_2011,cote_2014,andradibrown_2017}. The allometric
length-weight relationship is thus an essential component of these
models, but this relationship can vary across regions as a response to
biotic and abiotic conditions \citep{johnson_2016}.

@@

Our objective was to quantify the magnitude of error caused by using
\emph{ex situ} parameter values when estimating lionfish weight from
length observations. In this study, we calculated and reported the first
length-weight relationship for lionfish in the central Mexican Caribbean
using previously collected \emph{in situ} observations (n = 109;
\citep{villaseorderbez_2014}). We then estimated lionfish weight in this
area using previously published length-weight relationships for lionfish
populations from ten locations across the Western Atlantic, Gulf of
Mexico, and Caribbean. By comparing these weight estimates to our
\emph{in situ} length-weight observations, we showed that using \emph{ex
situ} parameter values resulted in up to a threefold under- or
overestimation of lionfish weight.

@@

where \(a\) is the ponderal index and \(b\) is the scaling exponent or
allometric parameter. We linearized the equation using
\(log_{10}\)-transformation and estimated the coefficients using an
Ordinary Least Squares Regression with a heteroskedastic-robust standard
error correction \citep{zeileis_2004}. Coefficients were tested with a
two-tailed Student's t-test, and the significance of the regression was
corroborated with an F-test. Some of the reviewed studies (Table
\ref{tab:all_params}, Fig \ref{fig:map}) inconsistently defined \(a\) as
either the ponderal index from Eq. \ref{eq:allometric} or the
y-intercept from the linearized log-transformed equation. Other studies
incorrectly reported parameters as mm-to-g conversions when they were in
fact cm-to-g conversions. We standardized each study by converting
coefficients and report all parameters as TL (mm) to TW (g) conversions.

We obtained a total of 18 parameter pairs by combining length-weight
parameters extracted from the literature and the additional pair
calculated here (Fig \ref{fig:all_allo}). Recall that the objective of
this study is not to describe variations between populations (which
would require access to raw data), but rather to assess how \emph{ex
situ} parameter values influence the accuracy of weight estimates for
lionfish, using the central Mexican Caribbean as a case study. Using
each of the 18 parameter pairs, we estimated \(TW\) from the \(TL\)
observations collected in the central Mexican Caribbean (n = 109, with
\(TL\) ranging from 34 mm to 310 mm) and divided predicted weights by
known observed weights to obtain a simple measure of over- or
underestimation. Difference in mean weight ratios were tested with an
analysis of covariance (ANCOVA):

@@

Where \(R_{ij}\) is the weight ratio for the \(i\)-th organism obtained
with parameters from the \(j\)-th study, \(\tilde{\mu}\) is a constant
for all individuals, \(a_j\) is the treatment effect (\emph{i.e.} the
difference induced by each study), \(TL_{ij}\) is the covariate
(\emph{i.e.} Total Length) for the \(i\)-th subject in the \(j\)-th
group with slope \(\beta\), and \(e_{ij}\) is the error term of the
regression. Ratios were logit-transformed prior to analysis, and a
\emph{post-hoc} Tukey's test was used to identify groups where mean
ratios did not differ. All analyses were performed in R version 3.5.2
\citep{rcore_2018}. Raw data and code used in this work are available on
github at github.com/jcvdav/lionfish\_biometry.

@@

The length-weight relationship for organisms from the central Mexican
Caribbean (\ref{fig:l-w-carib}) resulted in coefficient values of
\(a = 3.205 \times 10^{-6}\) and \(b = 3.235\) (\(R^2 = 0.977\),
\(F_{1, 107} = 6928.67; p < 0.001\)). The allometric factor (\(b\)) was
significantly different from \(b = 3\) (\(t_{107} = 6.04; p<0.001\)),
corroborating that lionfish present allometric growth. The length-weight
coefficients estimated here were within the range identified by studies
from other regions (Table \ref{tab:all_params}).

@@

ANCOVA results show that there were significant differences in our
predicted weight ratios for the central Mexican Caribbean when using
each of the different pairs of parameters
(\(F_{17, 1943} = 24.96; p < 0.001\); Fig \ref{fig:bio_ratio}). For
example, the actual observed weights of the 109 lionfish from the
Central Mexican Caribbean had a mean \(\pm\) SD of 52.56 \(\pm\) 76.58
g. However, if we used allometric parameter values from Banco Chinchorro
in the Caribbean to predict weights from our observed length
observations, we estimated a mean \(\pm\) SD of 40.37 \(\pm\) 58.74 g
\citep{sabidoitz_2016}. If we similarly used parameter values from North
Carolina in the Western Atlantic to estimate lionfish weights in the
Central Mexican Caribbean, we found a mean \(\pm\) SD 73.76 \(\pm\)
96.11 g \citep{barbour_2011}. Weights predicted from these extreme
parameters correspond to mean predicted-to-observed weight ratios of
0.80 \(\pm\) 0.19 and 1.76 \(\pm\) 0.50 (mean \(\pm\) SD), respectively.
Furthermore, for individual organisms, the largest differences between
predicted and observed values resulted in ratios of 0.36 and 3.51
(\emph{i.e.} the tails of each violin in Fig \ref{fig:bio_ratio}). If
instead of looking at the mean ratios we looked at the total biomass
(\emph{i.e.} summing across organisms), these cases would lead to
estimated total biomasses of 76.2\% (4,363.53 g) and 140\% (8,039.96 g)
of true observed biomass (5,729.34 g). These were the two most extreme
sets of parameters, but results consistently varie Use of \emph{ex situ}
parameters can result in erroneous weight and total biomass estimates
for lionfish.

Tukey's \emph{post-hoc} test suggested that weight ratios for the
central Mexican Caribbean were not different from those obtained with
parameters from Little Cayman, the Bahamas, and some sites in the Gulf
of Mexico (Tukey's HSD \(p > 0.05\)). All weight estimates using
parameters from the Gulf of Mexico and Western Atlantic were higher than
observed values, and only parameters from the Caribbean produced weights
smaller than observed (Fig \ref{fig:bio_ratio}). The regional average
(\(\pm\) SD) of predicted-to-observed weight ratios from these three
regions were 1.24 \(\pm\) 0.309, 1.41 \(\pm\) 0.523, and 1.20 \(\pm\)
0.423 for the Gulf of Mexico, Western Atlantic, and Caribbean,
respectively. This suggests that the smallest errors are observed when
using parameters from other locations in the Caribbean.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/pred_obs-1.pdf}
\caption{\label{fig:bio_ratio}Violin plot of predicted-to-observed
weight ratios for 18 pairs of allometric parameters. Sex is indicated in
parentheses. Blue circles indicate median values and like letters
indicate values that do not differ significantly. For Sabido-Itza et al,
2016b, BC and PNAX make reference to Banco Chinchorro and Parque
Nacional Arrecifes de Xcalak, two sites for which they report
parameters.}
\end{figure}

@@

Our results suggest that lionfish exhibit highly variable, spatially
heterogeneous allometric relationships across their invaded range, and
that this variation is relevant for managing invasions. Moreover, we
show that the use of \emph{ex situ} parameter values may lead to highly
biased weight estimates. Our comparison of observed weights to those
predicted with locally-informed parameters and \emph{ex situ} parameters
showed that weight can be overestimated by more than threefold,
highlighting the need to use local information. Here we discuss the
implications of our findings, possible shortcommings in our analyses,
and highlight potential future research directions.

Differences in length-weight relationships have traditionally been
highlighted as potential pitfalls to fishery management. For example,
\citet{wilson_2012} show that small-scale variations in length-at-age
and fishing mortality in other Scorpaeniformes translate to differential
landings, effort, and catch per unit effort in the live fish fishery of
California, and that these differences must be taken into account in
management plans. The lionfish case poses the opposite scenario, where
the manager desires to eradicate the species. To accurately gauge both
the effectiveness of lionfish removal efforts and the resources needed
to successfully manage an invasion, we must acknowledge and understand
regional biological differences in important variables such as
allometric growth parameters.

@@

We might be inclined to attribute all variation in the lionfish
length-weight relationship to the spatial origin of these parameters.
However, samples from the 12 studies included here were not only
collected in different locations, but also at different points in time
and across different depth and size ranges (See Supplementary Table 2
for an extended version of Table 1). While we could not evaluate how
these factors influenced length-weight estimates from previous studies
without raw data, we still show that a lack of locally-calculated
parameters can induce significant bias when calculating weight from
length observations. The magnitude of the bias discovered in this study
and our lack of understanding the sources driving spatial variation for
lionfish highlights the need to simultaneously collect length-weight
information across the invaded range to test for spatially-induced
patterns and to link these findings to previously suggested
environmental and genetic structures. Such an endeavor would provide
insight into lionfish biology and better inform management.

Applying parameter estimates to lengths outside the range of lengths
originally used to estimate the parameters may also induce error. Our
smallest observed organism was 34 mm in \(TL\), and only two studies
estimated parametrs with smaller organisms
\citep{edwards_2014,sabidoitza_2016}. By contrast, our largest organism
had a \(TL\) of 310 mm, which is well within the range of all other
studies (the next smallest maximum length was 325 mm; See Supplementary
Table 2). Due to the power function describing the allometric
relationship (\emph{i.e.} Eq. 1), the error in weight estimates is
larger when extrapolation is done for lengths that are larger than the
maximum length used to estimate the parameters. Our estimates are
therefore conservative because we only used parameter pairs from other
studies to estimate weights for lionfish up to 310 mm in the Central
Mexican Caribbean, well within the range of lengths for which other
parameters were estimated.

>>>>

\begin{abstract}
Lionfish (\textit{Pterois volitans / miles}) are an invasive species in
the Western Atlantic and the Caribbean. Improving management of invasive
lionfish populations requires accurate total biomass estimates, which
depend on accurate estimates of allometric growth, but sedentary species
like lionfish often exhibit high levels of spatial variation in life
history characteristics. We reviewed 17 published length-weight
relationships for lionfish taken throughout their invasive range and
found regional differences that led to significant misestimates when
calculating weight from length observations. The spatial pattern we
observed is consistent with findings from other studies focused on
genetics or length-at-age. Here, the use of \textit{ex situ} parameter
values resulted in total biomass estimates between 76.2\% and 140\% of
true observed biomass, and up to a threefold under- or overestimation of
total weight for an individual organism. These findings can have
implications for management in terms of predicting effects on local
ecosystems, evaluating the effectiveness of removal programs, or
estimating biomass available for harvest.
\end{abstract}

@@

Governments and non-profit organizations have sought to reduce lionfish
densities through removal programs and by incentivizing its consumption
\citep{chin_2016}. In some cases, these have shown to significantly
reduce -- but not quite eliminate -- lionfish abundances at local scales
\citep{deleon_2013,sandel_2015}. Complete eradication of lionfish
through fishing is unlikely because of their rapid recovery rates and
ongoing recruitment to shallow-water areas from persistent populations
in mesophotic ecosystems \citep{barbour_2011,andradibrown_2017}.
However, promoting lionfish consumption might create a level of demand
capable of incentivizing a stable fishery while controlling
shallow-water populations, thus creating alternative livelihoods and
avoiding further negative effects to local biota.

The feasibility of establishing fisheries through lionfish removal
programs has been extensively evaluated through field observations and
empirical modeling
\citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,usseglio_2017}.
Determining the feasibility of such initiatives requires modeling the
change in biomass in response to changes in fishing mortality
(\emph{i.e.,} culling). A common way to model this is via
length-structured population models, where fish lengths are converted to
weight to calculate total biomass
\citep{barbour_2011,cote_2014,andradibrown_2017}. The allometric
length-weight relationship is thus an essential component of these
models, but this relationship can vary across regions as a response to
biotic and abiotic conditions \citep{johnson_2016}.

@@

Our objective was to quantify the magnitude of error caused by using
\emph{ex situ} parameter values when estimating lionfish weight from
length observations. In this study, we calculated and reported the first
length-weight relationship for lionfish in the central Mexican Caribbean
using previously collected \emph{in situ} observations (n = 109;
\citet{villaseorderbez_2014}). We then estimated lionfish weight in this
area using previously published length-weight relationships for lionfish
populations from ten locations across the Western Atlantic, Gulf of
Mexico, and Caribbean. By comparing these weight estimates to our
\emph{in situ} length-weight observations, we showed that using \emph{ex
situ} parameter values resulted in up to a threefold under- or
overestimation of lionfish weight and estimated of total biomass ranged
between 76\% and 140\% of observed total biomass.

@@

Where \(a\) is the ponderal index and \(b\) is the scaling exponent or
allometric parameter. We linearized the equation using
\(log_{10}\)-transformation and estimated the coefficients using an
Ordinary Least Squares Regression with a heteroskedastic-robust standard
error correction \citep{zeileis_2004}. Coefficients were tested with a
two-tailed Student's t-test, and the significance of the regression was
corroborated with an F-test. Some of the reviewed studies (Table
\ref{tab:all_params}, Fig \ref{fig:map}) inconsistently defined \(a\) as
either the ponderal index from Eq. \ref{eq:allometric} or the
y-intercept from the linearized log-transformed equation. Other studies
incorrectly reported parameters as mm-to-g conversions when they were in
fact cm-to-g conversions. We standardized each study by converting
coefficients and report all parameters as TL (mm) to TW (g) conversions.

We obtained a total of 18 parameter pairs by combining length-weight
parameters extracted from the literature and the additional pair
calculated here (Fig \ref{fig:all_allo}). Recall that the objective of
this study is not to describe differences in the length-weight
relationship between populations (which would require access to raw
data), but rather to assess how \emph{ex situ} parameter values
influence the accuracy of weight estimates for lionfish, using the
central Mexican Caribbean as a case study. Using each of the 18
parameter pairs, we estimated \(TW\) from the \(TL\) observations
collected in the central Mexican Caribbean (n = 109, with \(TL\) ranging
from 34 mm to 310 mm) and divided predicted weights by known observed
weights to obtain a simple measure of over- or underestimation.
Difference in mean weight ratios were tested with an analysis of
covariance (ANCOVA):

@@

Where \(R_{ij}\) is the weight ratio for the \(i\)-th organism obtained
with parameters from the \(j\)-th study, \(\tilde{\mu}\) is a constant
for all individuals, \(a_j\) is the treatment effect (\emph{i.e.,} the
difference induced by each study), \(TL_{ij}\) is the covariate
(\emph{i.e.,} Total Length) for the \(i\)-th subject in the \(j\)-th
group with slope \(\beta\), and \(e_{ij}\) is the error term of the
regression. Ratios were logit-transformed prior to analysis, and a
\emph{post-hoc} Tukey's test was used to identify groups where mean
ratios did not differ. All analyses were performed in R version 3.5.2
\citep{rcore_2018}. Raw data and code used in this work are available on
github at github.com/jcvdav/lionfish\_biometry.

@@

The length-weight relationship for organisms from the central Mexican
Caribbean (Fig \ref{fig:l-w-carib}) resulted in coefficient values of
\(a = 3.205 \times 10^{-6}\) and \(b = 3.235\) (\(R^2 = 0.977\),
\(F_{1, 107} = 6928.67; p < 0.001\)). The allometric factor (\(b\)) was
significantly different from \(b = 3\) (\(t_{107} = 6.04; p<0.001\)),
corroborating that lionfish present allometric growth. The length-weight
coefficients estimated here were within the range identified by studies
from other regions (Table \ref{tab:all_params}).

@@

ANCOVA results revealed significant differences in our predicted weight
ratios for the central Mexican Caribbean when using each of the
different pairs of parameters (\(F_{17, 1943} = 24.96; p < 0.001\); Fig
\ref{fig:bio_ratio}). For example, the actual observed weights of the
109 lionfish from the Central Mexican Caribbean had a mean \(\pm\) SD of
52.56 \(\pm\) 76.58 g. However, if we used allometric parameter values
from Banco Chinchorro in the Caribbean to predict weights from our
observed length observations, we estimated a mean \(\pm\) SD of 40.37
\(\pm\) 58.74 g \citep{sabidoitz_2016}. If we similarly used parameter
values from North Carolina in the Western Atlantic to estimate lionfish
weights in the Central Mexican Caribbean, we found a mean \(\pm\) SD of
73.76 \(\pm\) 96.11 g \citep{barbour_2011}. Weights predicted from these
extreme parameters correspond to mean predicted-to-observed weight
ratios of 0.80 \(\pm\) 0.19 and 1.76 \(\pm\) 0.50 (mean \(\pm\) SD),
respectively. Furthermore, largest errors for individual organisms
collected in the central Mexican Caribbean resulted in ratios of 0.36
and 3.51 (\emph{i.e.,} the tails of each violin in Fig
\ref{fig:bio_ratio}). If we examined biomass (\emph{i.e.,} summing
across all 109 organisms) instead of mean ratios, total biomass
estimates were 76.2\% (4,363.53 g) and 140\% (8,039.96 g) of true
observed biomass (5,729.34 g). Parameters for this study estimate total
biomass at 98\% of observed biomass. These misestimates come from the
two most extreme sets of parameters, but results varied consistently
across locations (Figs \ref{fig:bio_ratio} and \ref{fig:errors}).
Overall, the use of \emph{ex situ} parameters led to significantly
erroneous estimates of individual weight and total biomass for lionfish.

Tukey's \emph{post-hoc} test showed that weight ratios for the central
Mexican Caribbean differeed from those obtained with parameters from the
Western Atlantic, and most sites in the Caribbean and the Gulf of Mexico
(Tukey's HSD \(p > 0.05\)). The only sites where weight ratios did not
differ from the central Mexican Caribbean were Little Cayman
\citep{edwards_2014}, Bahamas \citep{darling_2011},
and the Northern Gulf of Mexico (\citet{dahl_2014}; Tukey's HSD
\(p > 0.05\)). All weight estimates using parameters from the Gulf of
Mexico and Western Atlantic were higher than observed values, and only
parameters from the Caribbean produced weights smaller than observed
(Fig \ref{fig:bio_ratio}). The regional average (\(\pm\) SD) of
predicted-to-observed weight ratios from these three regions were 1.24
\(\pm\) 0.309, 1.41 \(\pm\) 0.523, and 1.20 \(\pm\) 0.423 for the Gulf
of Mexico, Western Atlantic, and Caribbean, respectively. This suggests
that the smallest errors are observed when using parameters from other
locations in the Caribbean.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/pred_obs-1.pdf}
\caption{\label{fig:bio_ratio}Violin plot of predicted-to-observed
weight ratios when applying each of 18 different pairs of allometric
parameters to the 109 lionfish collected in the central Mexican
Caribbean. Sex is indicated in parentheses. Blue circles indicate median
values and like letters indicate values that do not differ
significantly. For Sabido-Itza et al, 2016b, BC and PNAX make reference
to Banco Chinchorro and Parque Nacional Arrecifes de Xcalak, two sites
for which they report parameters.}
\end{figure}

@@

Our results suggest that lionfish exhibit highly variable, spatially
heterogeneous allometric relationships across their invaded range, and
that this variation is relevant for managing invasions. Moreover, we
show that the use of \emph{ex situ} parameter values may lead to highly
biased weight and total biomass estimates. Our comparison of observed
weights to those predicted with locally-informed parameters and \emph{ex
situ} parameters showed that weight of an individual lionfish can be
overestimated by more than threefold, highlighting the need to use local
information. Here we discuss the implications of our findings, possible
shortcommings in our analyses, and highlight potential future research
directions.

Differences in length-weight relationships have traditionally been
highlighted as potential pitfalls to fishery management. For example,
\citet{wilson_2012} showed that small-scale variations in length-at-age
and fishing mortality in other Scorpaeniformes translate to differential
landings, effort, and catch per unit effort in the live fish fishery of
California, and that these differences must be taken into account in
management plans. The lionfish case poses the opposite scenario, where
the manager desires to eradicate the species. To accurately gauge both
the effectiveness of lionfish removal efforts and the resources needed
to successfully manage an invasion, we must acknowledge and understand
regional biological differences in important variables such as
allometric growth parameters.

@@

One might be inclined to attribute all variation in the lionfish
length-weight relationship to the spatial origin of these parameters.
However, samples from the 12 studies included here were not only
collected in different locations, but also at different points in time
and across different depth and size ranges (See Supplementary Table 2
for an extended version of Table 1). The magnitude of the bias
discovered in this study and our lack of understanding the sources
driving spatial variation for lionfish highlights the need to
simultaneously collect length-weight information across the invaded
range to test for spatially-induced patterns and to link these findings
to previously suggested environmental and genetic structures. Such an
endeavor would provide insight into lionfish biology and better inform
management. However, while we could not evaluate how these factors
influenced length-weight estimates from previous studies without raw
data, we still show that a lack of locally-calculated parameters can
induce significant bias when calculating weight from length
observations. We demonstrate the importance of using \emph{in situ}
parameters to obtain accurate weight estimates regardless of the
underlying mechanisms driving variation between populations.

Applying parameter estimates to lengths outside the range of lengths
originally used to estimate the parameters may also induce error. Our
smallest observed organism was 34 mm in \(TL\), and only two studies
estimated parametrs with smaller organisms
\citep{edwards_2014,sabidoitza_2016}. By contrast, our largest organism
had a \(TL\) of 310 mm, which is well within the range of all other
studies (the next smallest maximum length was 325 mm; See Supplementary
Table 2). Due to the power function describing the allometric
relationship (\emph{i.e.,} Eq. 1), the error in weight estimates is
larger when extrapolation is done for lengths that are larger than the
maximum length used to estimate the parameters. Our estimates are
therefore conservative because we only used parameter pairs from other
studies to estimate weights for lionfish up to 310 mm in the Central
Mexican Caribbean, well within the range of lengths for which other
parameters were estimated.
